Beyond ``Horizontal'' and ``Vertical'':\\ The Welfare Effects of Complex Integration\tnoteref{t1}

### 2024-08-29 doc/VerticalArrangements.tex
peggy's rewrite of the republic/santek merger section. copy emailed to charles on 20240829

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Empirical Application: Republic/Santek \label{sec:application}}

@@

Republic Services, the second largest waste management company in the U.S., acquired Santek Environmental in 2021.\footnote{See Competitive Impact Statement: U.S. and State of Alabama v. Republic Services, Inc.and Santek Waste Services, LLC, \url{https://www.justice.gov/atr/case-document/file/1382626/download}.}  The DOJ filed a complaint and negotiated a settlement in the case, alleging horizontal anti-competitive effects for four small-container commercial waste (SCCW) collection markets and two municipal solid waste (MSW) disposal markets. In addition, vertical anti-competitive effects were alleged to arise from the combination of Republic and Santek integrated assets in the Chattanooga region.\footnote{See U.S. and State of Alabama v. Republic Services, Inc. and Santek Waste Services, LLC, \url{https://www.justice.gov/atr/case-document/file/1382031/download}.}  In this section, we apply our model to simulate the effects of the merger in Chattanooga.\footnote{The \href{https://CRAN.R-project.org/package=antitrust}{\texttt{antitrust} R package} contains the computer code needed to run the merger simulations described here.}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Merger Simulation Results}

Overall, the merger between Republic and Santek is estimated to lower upstream prices by 2.8\% and increase downstream prices by 12\%, resulting in \$16 million of annual harm to consumers and total annual producer benefits of approximately \$13.4 million. Thus, on net the model predicts the merger would be harmful despite the presence of significant EDM that reduces prices upstream.\footnote{Our calibrated bargaining parameters for retailers against Santek ($\lambda=1$) and Republic ($\lambda=0.7$) give a substantial edge to downstream firms, which, based on our analysis of the previous section, suggests consumers are likely to be harmed.  High bargaining power parameters are needed to explain the high collection margin compared to the disposal margin.  (See Table \ref{tab:trashdata}.)}  These effects are illustrated in more detail in Figure \ref{fig:TrashSimsFirms}.

Disposal prices for pre-merger integrated pairs are unchanged post-merger. However, MSW disposal prices increase to all rival haulers post-merger (i.e., all unintegrated pairs). The merged firms' disposal prices to downstream rivals are estimated to increase 88\% post-merger, and Waste Connections' prices to its downstream rivals are estimated to increase 22\%. These large price increases are offset by the large decrease in the disposal price for the integrating pair post-merger due to EDM, as the cost to Republic for volume disposed with Santek decreases by 60\%.

Turning to the downstream market, all collection prices increase post-merger. The merging parties' post-merger collection prices increase by approximately 20\%, and their share decreases by about 8 percentage points overall. The Republic/Republic product's price increases 21\%, and the Santek/Santek product's price increases by 23\%. The newly integrated Santek/Republic product's price is estimated to increase by 9\%, demonstrating that EDM is not fully passed-through to consumers. The integrated rival, Waste Connections, increases its collection price by approximately 5\%, and its share increases by 6 percentage points. The unintegrated rivals' collection prices each increase by about 9\%, which again demonstrates incomplete pass-through of the change in upstream costs, and their shares increase slightly.

@@

Ignoring these vertical relationships in the downstream-only model results in an estimated 11\% increase in collection prices with a greater than 16\% increase in the merging parties' price, consumer harm of \$10 million annually, and net harm of approximately \$2 million. Estimated consumer harm is 60\% higher in the full vertical model and is offset by markedly larger producer benefits. The downstream-only model misses the substantial RRC effect experienced by the Santek/WM-ADS and Santek/Regional products.  In the vertical model, the downstream prices for these products rise more dramatically due to this RRC, and other prices increase alongside them.  This causes more harm to consumers.

In the upstream-only model, downstream prices and therefore shares are fixed at their pre-merger levels. We still have upstream prices set to marginal cost for integrated products.  Overall disposal prices increase by 8.2\% post-merger, and the merger effect is a pure transfer between firms. Thus, both consumer and net harm are estimated to be zero.  The main impact of the merger is in changing the bargaining leverage for Santek, now that it is combined with Republic.  This effect causes Santek to raise its upstream disposal prices to both WM-ADS and Regional.  The price increase is even larger in the upstream-only model compared to the full model. The lack of downstream price adjustments means that WM-ADS and Regional cannot shift their sales towards their cheaper supplier, Waste Connections, in equilibrium, which makes having access to Santek's disposal facilities relatively more valuable than in the full model, driving up Santek's price.  In this sense, the upstream-only model overstates the increase in leverage that Santek receives from the merger.

@@

@@

\begin{figure}
\centering
\caption{Republic/Santek Simulations: Full vs. Partial Models}
\includegraphics[scale=1.0]{output/TrashSimsCompareAll.png}\\
\footnotesize{Notes: ``WasteConn'' stands for Waste Connections.}
\label{fig:TrashSimsFirms}
\end{figure}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Empirical Application: the Chattanooga Waste Industry \label{sec:application}}

@@

Republic Services, the second largest waste management company in the U.S., acquired Santek Environmental in 2021.\footnote{See Competitive Impact Statement: U.S. and State of Alabama v. Republic Services, Inc.and Santek Waste Services, LLC, \url{https://www.justice.gov/atr/case-document/file/1382626/download}.}  The DOJ filed a complaint and negotiated a settlement in the case, alleging horizontal anti-competitive effects for four small-container commercial waste (SCCW) collection markets and two municipal solid waste (MSW) disposal markets. In addition, vertical anti-competitive effects were alleged to arise from the combination of Republic and Santek integrated assets in the Chattanooga region.\footnote{See U.S. and State of Alabama v. Republic Services, Inc. and Santek Waste Services, LLC, \url{https://www.justice.gov/atr/case-document/file/1382031/download}.}  In this section, we begin by applying our model to simulate the effects of the Republic/Santek merger in Chattanooga. We then compare these results to those of counterfactual merger simulations in which we examine how competition might have been affected by the combination of other market competitors that exhibit different vertical supply chain arrangements.\footnote{The \href{https://CRAN.R-project.org/package=antitrust}{\texttt{antitrust} R package} contains the computer code needed to run the merger simulations described here.} By focusing on one market at a single point in time, we are better able to interrogate the welfare implications of these structural differences.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Republic/Santek Merger Simulation Results}

Overall, the merger between Republic and Santek is estimated to lower upstream prices by \hl{2.8\%} and increase downstream prices by \hl{12\%}, resulting in \$16.6 million of annual harm to consumers and total annual producer benefits of approximately \$14 million. Thus, on net the model predicts the merger would be harmful despite the presence of significant EDM that reduces prices upstream.\footnote{Our calibrated bargaining parameters for retailers against Santek ($\lambda=1$) and Republic ($\lambda=0.7$) give a substantial edge to downstream firms, which, based on our analysis of the previous section, suggests consumers are likely to be harmed.  High bargaining power parameters are needed to explain the high collection margin compared to the disposal margin.  (See Table \ref{tab:trashdata}.)}  These effects are illustrated in more detail in
Figure \ref{fig:TrashSimsFirms}.

Disposal prices for pre-merger integrated pairs are unchanged post-merger. However, MSW disposal prices increase to all rival haulers post-merger (i.e., all unintegrated pairs). The merged firms' disposal prices to downstream rivals are estimated to increase over 91\% post-merger, and Waste Connections' prices to its downstream rivals are estimated to increase 24\%. These large price increases are offset by the large decrease in the disposal price for the integrating pair post-merger due to EDM, as the cost to Republic for volume disposed with Santek decreases by approximately 61\%.

Turning to the downstream market, all collection prices increase post-merger. The merging parties' post-merger collection prices increase by approximately \hl{20\%}, and their share decreases by about 8 percentage points overall. The Republic/Republic product's price increases 21\%, and the Santek/Santek product's price increases by 23\%. The newly integrated Santek/Republic product's price is estimated to increase by 9\%, demonstrating that EDM is not fully passed-through to consumers. The integrated rival, Waste Connections, increases its collection price by approximately 6\%, and its share increases by 6 percentage points. The unintegrated rivals' collection prices each increase by about \hl{9\%}, which again demonstrates incomplete pass-through of the change in upstream costs, and their shares increase slightly.

@@

Ignoring these vertical relationships in the downstream-only model results in an estimated
\hl{11\%} increase in collection prices with a greater than \hl{16\%} increase in the merging parties' price, consumer harm of \hl{\$10 million} annually, and net harm of approximately \hl{\$2 million}. Estimated consumer harm is \hl{60\%} higher in the full vertical model and is offset by markedly larger producer benefits. The downstream-only model misses the substantial RRC effect experienced by the Santek/WM-ADS and Santek/Regional products.  In the vertical model, the downstream prices for these products rise more dramatically due to this RRC, and other prices increase alongside them.  This causes more harm to consumers.

In the upstream-only model, downstream prices and therefore shares are fixed at their pre-merger levels. Upstream prices are still set to marginal cost for integrated products.  Overall disposal prices increase by \hl{8.2\%} post-merger, and the merger effect is a pure transfer between firms. Thus, both \hl{consumer and net harm are estimated to be zero}.  The main impact of the merger is in changing the bargaining leverage for Santek, now that it is combined with Republic.  This effect causes Santek to raise its upstream disposal prices to both WM-ADS and Regional.  The price increase is even larger in the upstream-only model compared to the full model. The lack of downstream price adjustments means that WM-ADS and Regional cannot shift their sales towards their cheaper supplier, Waste Connections, in equilibrium, which makes having access to Santek's disposal facilities relatively more valuable than in the full model, driving up Santek's price.  In this sense, the upstream-only model overstates the increase in leverage that Santek receives from the merger.

@@

\subsection{Counterfactual Merger Simulations}

In addition to applying our model to estimate the effects of the Republic/Santek merger on the Chattanooga market, in this section we simulate the effects of the hypothetical mergers of three other pairs (Santek/Waste Connections, Santek/WM-ADS, and WM-ADS/Regional) that would have resulted in companies with different vertical supply configurations combining. The estimated aggregate welfare effects of these counterfactual mergers are illustrated alongside those of the Reublic/Santek merger in Figure \ref{fig:TrashFakeOverall}.

The pre-merger Herfindahl-Hirschman Index (HHI) for the Chattanooga disposal market is \hl{xxxx} and collection market is \hl{yyyy}. The merger of WM-ADS with the Regional collection firm is a straightforward ``horizontal'' merger between two relatively small downstream competitors. This merger results in a post-merger change in the collection market HHI of \hl{zzz}. Unsurprisingly, the resulting estimated welfare effects are small and consumer harm and producer benefits approximately cancel each other out. The more detailed results reported in Figure \ref{fig:TrashFakeFirms} show only minimal changes in prices or collection shares post-merger.

Turning to the Santek/WM-ADS merger simulation, we consider the effects of a small downstream rival combining with a vertically integrated firm. Here the merger results in a post-merger change in the collection market HHI of \hl{aaa}, but consumer harm from increased collection prices of \hl{x\%} is significantly dampened by price decreases in collection of \hl{y\%} from EDM. This results in net harm of approximately \hl{\$z million}.

Finally, we examine the potential merger of Santek and Waste Connections. While this merger is similar to that of Santek and Republic in that it is a merger between veritcally integratted rivals, Santek/WasteConn differs in it's effects on the Chattanooga market because both parties exclusively self-supplied for their disposal input pre-merger. This diminishes the possibility for EDM effects. In addition, while Santek/WasteConn significantly raise disposal prices to their downstream rivals, the presence of the larger, vertically integrated (even if capacity constrained) Republic limits the merged firm's ability to pass this through to the collection market. Thus, net harm to the Chattanooga market remains relatively small at \hl{\$d million}. This is in spite of post-merger changes in the HHIs of the disposal market of \hl{xxxx} and collection market of \hl{yyyy}.

\subsection{Robustness}

The robustness of our simulation results depends critically on a few key parameter values. One of these is the assumption of the bargaining power parameter, $\lambda$. Sheu and Taragin (2021) shows that mergers tend to benefit consumers when wholesalers have relatively more power, and harm consumers when retailers have relatively more power. Larger wholesaler bargaining power offers more possibilities for EDM, as premerger input prices are likely to be high. This channel becomes less relevant as retailers gain more power. Figure \ref{fig:TrashBarg} illustrates his relationship between consumer welfare and the value of the bargaining power parameter for the Republic/Santek merger simulation. Harm increases rapidly as bargaining power increases to $0.4$ and then becomes nearly constant for values above $0.5$. Thus, assuming an equal split between collection and disposal would yield an estimate of a near upper bound on consumer harm for this merger.

In addition to specification of the bargaining power parameter, our results are highly sensitive to assumptions about the market elasticity which is specified through the share of consumers choosing to purchase goods outside the market (the outside good share). Our baseline results assume that the outside share is zero because there are essentially no reasonable substitutes for SCCW collection or MSW disposal.\footnote{See Competitive Impact Statement: U.S and Plaintiff States v. Republic Services, Inc. and Santek, LLC, \url{https://www.justice.gov/atr/case-document/file/1382626/dl?inline}} However, this assumption leads to the largest possible consumer harm in our model as shown in Figure \ref{fig:TrashElast}. Often when the size of the outside share is unknown a market elasticity of 1.0 is assumed. For the Republic/Santek merger example, this would lead to consumer harm approximately 10 percentage points smaller.

@@

\begin{figure}
\centering
\caption{Republic/Santek Simulations: Full vs. Partial Models}
\includegraphics[scale=1.0]{output/TrashSimsCompareAll.png}\\
\footnotesize{Notes: ``WasteConn'' stands for Waste Connections.}
\label{fig:TrashSimsCompareAll}
\end{figure}
